Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-17 paper.tex
Define matches as lists

A \textit{match} $M$ for a given rule $r$ is a partial map from slots of $r$ to hypotheses.
The map must satisfy the following conditions:
\begin{enumerate}
  \item\label{cond:match:matches} For each slot $i ∈ \dom(M)$, the hypothesis $M(i)$ matches $i$.
  \item\label{cond:match:compat} For all slots $i,j ∈ \dom(M)$, the substitutions $\sub_{i}(M(i))$ and $\sub_{j}(M(j))$ are \emph{compatible}.
        This means that for each variable $x ∈ \dom(\sub_{i}(M(i))) ∩ \dom(\sub_{j}(M(j)))$ we have $\sub_{i}(M(i))(x) = \sub_{j}(M(j))(x)$, i.e.\ the substitutions agree on all common variables.
  \item\label{cond:match:closed} $M$ is \emph{downward-closed}: if $i ∈ \dom(M)$, then for any slot $j < i$ of $r$ we also have $j ∈ \dom(M)$.
\end{enumerate}
The \emph{substitution of a match $M$}, $\sub(M)$, is the union of the substitutions induced by the hypotheses in $M$, i.e. $\sub(M) ≔ ⋃_{i ∈ \dom(M)}\sub_{i}(M(i))$.
This is well-defined by requirement~\ref{cond:match:compat}.
The \textit{level of a match $M$}, $\lvl(M)$, is the maximal $i$ such that $i ∈ \dom(M)$.
A match for rule $r$ is \textit{complete} if $i ∈ \dom(M)$ for every slot $i$ of $r$, and \emph{partial} otherwise.

A partial match $M$ for rule $r$ corresponds to a partial application of the rule to the terms given in $\sub(M)$.
The application is type-correct by requirements~\ref{cond:match:matches} and~\ref{cond:match:compat}.
If $M$ is complete, then $\sub(M)$ contains a term for every slot and every variable---hence for every premise---of $r$ and we have a full application.

Requirement~\ref{cond:match:closed} implies that we will be adding hypotheses to matches in the order in which the slots appear.
Hence for a rule $r : A → B → C → D$, we effectively consider partial applications $r~a$ and $r~a~b$ for some $a : A$ and $b : B$, but not $r~\text{--}~b$.
This prevents us from constructing $r~a~b$ in two different but ultimately redundant ways.
The term orders used in various saturation calculi serve a similar function.

@@

Prior to this work, Aesop used what we call naive forward reasoning (TODO ref).
This approach treats forward rules essentially like any other Aesop rule.
When Aesop selects a forward rule $r : \All{(x₁ : A₁) \dots (xₘ : Aₘ)}{B}$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we first unify the last premise $Aₘ$ with each hypothesis $h : C$ such that $C$ and $Aₘ$ likely unify (as determined by a discrimination tree (TODO ref) index).
If successful, this results in a substitution $σ₁$ containing $xₘ$ and any variables occurring in $Aₘ$.
We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$ (skipping premises $A_{m-i}$ if they are already determined by $σᵢ$).
If a premise $A_{m-i}[σᵢ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
These may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
If this process yields at least one full application $r~a₁ \dots aₘ$, we can add a new hypothesis of type $T = B[x₁ ↦ a₁, \dots, xₘ ↦ aₘ]$, meaning we generate the subgoal $Γ, h : T$ (unless $Γ$ already contains a hypothesis of type $T$).

@@

\paragraph{Forward State.}
We associate to each goal in Aesop's search tree a forward state that reflects the goal's context.
Let $\States$ be the set of forward states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
The forward state's interface then consists of two functions
\begin{equation*}
  \addHyp : \States → \Hyps → \States × \powerset(\Rules × \CMatches) \qquad \delHyp : \States → \Hyps → \States
\end{equation*}
The $\addHyp$ function adds a hypothesis to the forward state.
Hence, if $s$ is the forward state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', M)$, then $s'$ is the forward state corresponding to $Γ,\, h : T$.
Similarly, if $s$ is the forward state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the forward state corresponding to $Γ$.
The second output of $\addHyp$, $M$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
These are the matches that were completed by adding $h : T$.

@@

When a hypothesis is added to a rule state, two things can happen.
First, the hypothesis may allow us to extend partial matches already present in the state.
For instance, consider the rule $r : \All{x}{P~x → Q~x → R~x → S~x}$ and a rule state containing the partial match $\{1 ↦ h₁\}$, with $1$ referring to the slot of type $P~x$ and $h₁ : P~a$ for some $a$.
Adding a new hypothesis $h₂ : Q~a$ to this rule state means that we should add to the state the extended match $\{1 ↦ h₁, 2 ↦ h₂\}$.

Second, adding a match may trigger the insertion of further matches.
Suppose that before adding $h₁$ and $h₂$, we had already added a hypothesis $h₀ : R~a$ to the rule state.
Once the partial match $\{1 ↦ h₁, 2 ↦ h₂\}$ arrives, we should recognise that it can be extended with $h₀$ to obtain the (complete) match $\{1 ↦ h₁, 2 ↦ h₂, 3 ↦ h₀\}$.

\paragraph{Variable Indices.}
Extending the matches above was possible only because all hypotheses are compatible on the shared variable $x$, i.e.\ they all instantiate $x$ with the same term $a$.
In general, to extend a match $M$ with a hypothesis $h : T$ at slot $i$, we need to ensure that the substitution of $M$, $\sub(M)$, is compatible with the substitution of $h$, $\sub_{i}(h)$.

@@

\begin{algorithm}
  \SetKwInOut{Precondition}{Precondition}
  \SetKwInOut{Postcondition}{Postcondition}
  \Precondition{$H$ and $M$ contain the hypotheses and matches corresponding to a context $Γ$}
  \Input{A hypothesis $h : T$ matching slot $i$ of $r$ with substitution $σ$}
  \Output{Complete matches of $r$ in context $Γ,\, h : T$ that contain $h$}
  \Postcondition{$H$ and $M$ contain the hypotheses and matches corresponding to $Γ,\, h : T$}
  \BlankLine
  \Procedure{$\mathrm{addHyp}(h : T)$}{
    \eIf{$i = 1$}{
      $\mathrm{addMatch}(\{1 ↦ h\})$
    }{
      \For{$x$ \KwIn $\vars(i)$}{
        Insert $h$ into the set $H_{x}(i, σ(h))$
      }
      \KwLet $\mathit{ms} ≔ \bigcap_{x \in \vars(i)} M_{x}(i-1, σ(x))$ \\
      $\mathit{completeMs} \gets ∅$ \\
      \For{$m$ \KwIn $\mathit{ms}$}{
        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ∪ \{i ↦ h\})$
      }
      \Return{$\mathit{completeMs}$}
    }
  }
  \BlankLine % without these, the caption is not offset properly
  \BlankLine
  \caption{Adding a hypothesis to the rule state for a rule $r$}%
  \label{alg:addhyp}
\end{algorithm}

The rule state's $\addHyp$ method (Algorithm~\ref{alg:addhyp}) adds a hypothesis $h : T$ matching slot $i$ to the rule state.
(If a hypothesis matches multiple slots, $\addHyp$ is called multiple times.)
For $i = 1$, we add the singleton match $\{1 ↦ h\}$ to the rule state.
For $i > 1$, we first insert $h$ into the hypothesis index $H$.
Then we query the match index $M$ for matches that can be extended by $h$ and add the extended matches to the rule state.
While adding matches, we may discover new complete matches, which are collected and returned.

% TODO pre- and postcondition?
\begin{algorithm}
  \Input{A partial match $m$ of $r$ at level $i$ with substitution $σ$}
  \Output{A set of complete matches of $r$}
  \BlankLine
  \Procedure{$\mathrm{addMatch}(h : T)$}{
    \eIf{$i$ is complete}{
      \Return{$\{m\}$}
    }{
      \For{$x$ \KwIn $\vars(i + 1)$}{
        Insert $m$ into the set $M_{x}(i, σ(x))$
      }
      \KwLet $\mathit{hs} ≔ \bigcap_{x \in \vars(i + 1)} H_{x}(i + 1, σ(x))$ \\
      $\mathit{completeMs} \gets ∅$ \\
      \For{$h$ \KwIn $\mathit{hs}$}{
        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ∪ \{i + 1 ↦ h\})$
      }
      \Return{$\mathit{completeMs}$}
    }
  }
  \BlankLine % without these, the caption is not offset properly
  \BlankLine
  \caption{Adding a match to the rule state for a rule $r$}%
  \label{alg:addmatch}
\end{algorithm}

@@

\paragraph{Variable Clusters.}
We have so far assumed that for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
This assumption is violated for rules such as $\All{x~y}{P~x → Q~y → R~x~y}$: slot 2 ($Q~y$) shares no variables with slot 1 ($P~x$).
As a result, our above algorithms fail.
Suppose we are adding the hypothesis $h₂ : Q~b$, having already added the hypothesis $h₁ : P~a$ and hence the partial match $\{1 ↦ h₁\}$.
The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M_{x}(i-1, σ(x)) = ∅$, but the match $\{1 ↦ h₁\}$ can be extended with $h₂$.
In fact, any match at level 1 can be extended with any hypothesis matching slot 2 since the two slots are independent.

@@

\paragraph{Example 1.}
As a first example, consider a rule state for the transitivity rule $r : \All{x~y~z}{x ≤ y → y ≤ z → x ≤ z}$ to which we add, in order, the hypotheses
$h₁ : 301 ≤ 302$ and $h₂ : 300 ≤ 301$.
The rule has two slots of types $x ≤ y$ and $y ≤ z$ which share the variable $y$.

The hypothesis $h₁ : 301 ≤ 302$ unifies with both slots, so we run Algorithm~\ref{alg:addhyp} twice.
For slot 1, unification yields the substitution $\{x ↦ 301, y ↦ 302\}$.
Since this is the first slot, we start a new partial match $\{1 ↦ h₁\}$, which Algorithm~\ref{alg:addmatch} inserts into the set $M_{y}(1, 302)$.
This match cannot be extended since the hypothesis index $H$ does not contain a hypothesis for slot 2.

For slot 2, unification yields the substitution $σ ≔ \{y ↦ 301, z ↦ 302\}$, so we insert $h₁$ into $H_{y}(2, 301)$.
No matches at level 1 can be extended by $h₁$ since
\[
  \bigcap_{x \in \vars(2)} M_{x}(1, σ(x)) = M_{y}(1, 301) = \emptyset
\]
At this point, $H$ and $M$ are in the following state.
For brevity, we write matches $\{1 ↦ h₁, \dots, n ↦ hₙ\}$ as $[h₁, \dots, hₙ]$.
\begin{center}
\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
  \node[xshift = 0] {$H_{y}$}
    child {node {2}
      child {node {$301$}
        child {node {$\{h₁\}$}}}
    %  child {node {$300$}
    %    child {node {$\{\}$}}}
    };
  \node[xshift=160] {$M_y$}
    child {node {1}
      child {node {$302$}
        child {node {$\{[h₁]\}$}}}
    %  child {node {$301$}
    %    child {node {$\{\}$}}}
    };
    %child {node[xshift=20] {2}
    %  child {node {$301$}
    %    child {node[yshift=-2em] {$\{\}$}}}};
\end{tikzpicture}
\end{center}

Now we add $h₂ : 300 ≤ 301$, which again unifies with both slots.
For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $\{1 ↦ h₂\}$ to $M_{y}(1, 301)$.
We then check for hypotheses in slot 2 that also instantiate $y$ with $301$, and indeed we find $M_{y}(2, 301) = \{h₁\}$.
Hence we return the complete match $\{1 ↦ h₂, 2 ↦ h₁\}$, which induces a new hypothesis $h₃ : 300 ≤ 302$ to be added later.

@@

We now consider a rule $s : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $s$, starting with $h₁$:

@@

We now add $h_2 : A~a~b$, which initially generates a match $m = \{1 ↦ h₂\}$ that is added to $M$:

@@

\paragraph{Match Equivalence.}
Naively, two matches $\{1 ↦ h₁, \dots, n ↦ hₙ\}$ and $\{1 ↦ h₁', \dots, n ↦ hₙ'\}$ would be considered equal if they are structurally equal, i.e.\ $h₁ = h₁'$, \dots, $hₙ = hₙ'$.
However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
The complete matches $m₁ ≔ \{1 ↦ h₁, 2 ↦ h₃\}$ and $m₁' ≔ \{1 ↦ h₁, 2 ↦ h₄\}$ are structurally different, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
Generating both $m₁$ and $m₁'$ would therefore be redundant.
Similarly, the partial matches $m₂ ≔ \{1 ↦ h₁\}$ and $m₂' ≔ \{1 ↦ h₂\}$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete $m₂$ and $m₂'$.
Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each variable $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.
To ensure that we respect definitional equality, the instantiations in the substitutions are kept in RPINF.

@@

Given a set of propositions $\mathcal{P} := \{P_{i,j} | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i} | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : P_{i,1}\to \dots P_{i,6} \to Q_i$.
The procedures are then run on a context containing exactly $\mathcal{P}$.
In this case, we sketch the behavior of the stateless algorithm:
It pick a first hypothesis, $h_{i,j} : P_{i,j}$.
If $j < 6$, the index returns nothing and it selects another hypothesis.
For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
This gives a complexity of about $O(n^2)$.

>>>>

A \textit{match} $m$ for a give rule $r$ with $n$ slots is a list containing at least one and at most $n$ hypotheses.
We interpret $m$ as a partial map from slots of $r$ to hypotheses, so $\dom(m) = \{i \mid 1 ≤ i ≤ |m| \}$ and for each slot $i ∈ \dom(m)$, $m(i)$ is the hypothesis assigned to $i$.
Each match $m$ must satisfy the following requirements:
\begin{enumerate}
  \item\label{cond:match:matches} For each slot $i ∈ \dom(m)$, the hypothesis $m(i)$ matches $i$.
  \item\label{cond:match:compat} For all slots $i,j ∈ \dom(m)$, the substitutions $\sub_{i}(m(i))$ and $\sub_{j}(m(j))$ are \emph{compatible}.
        This means that the substitutions agree on all common variables, i.e.\ for each variable $x ∈ \dom(\sub_{i}(m(i))) ∩ \dom(\sub_{j}(m(j)))$ we have $\sub_{i}(m(i))(x) = \sub_{j}(m(j))(x)$.
\end{enumerate}
The \emph{substitution of a match $m$}, $\sub(m)$, is the union of the substitutions induced by the hypotheses in $m$, i.e. $\sub(m) ≔ ⋃_{i ∈ \dom(m)}\sub_{i}(m(i))$.
This is well-defined by requirement~\ref{cond:match:compat}.
The \textit{level of a match $m$}, $\lvl(m)$, is the length of $m$.
A match for rule $r$ is \textit{complete} if $i ∈ \dom(M)$ for every slot $i$ of $r$, and \emph{partial} otherwise.

A partial match $m$ for rule $r$ corresponds to a partial application of the rule to the terms given in $\sub(M)$.
The application is type-correct by requirements~\ref{cond:match:matches} and~\ref{cond:match:compat}.
If $m$ is complete, then $\sub(m)$ contains a term for every slot and every variable---hence for every premise---of $r$ and we have a full application.

Note that matches represent only those partial applications where arguments are added without gaps.
Hence for a rule $r : A → B → C → D$, the partial applications $r~a$ and $r~a~b$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\text{--}~b$.
This will prevent us from constructing $r~a~b$ in two different but ultimately redundant ways.
The term orders used in various saturation calculi serve a similar function.

@@

\jcom{Formulate this in terms of matches.}
Prior to this work, Aesop used what we call naive forward reasoning (TODO ref).
This approach treats forward rules essentially like any other Aesop rule.
When Aesop selects a forward rule $r : \All{(x₁ : A₁) \dots (xₘ : Aₘ)}{B}$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we first unify the last premise $Aₘ$ with each hypothesis $h : C$ such that $C$ and $Aₘ$ likely unify (as determined by a discrimination tree (TODO ref) index).
If successful, this results in a substitution $σ₁$ containing $xₘ$ and any variables occurring in $Aₘ$.
We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$ (skipping premises $A_{m-i}$ if they are already determined by $σᵢ$).
If a premise $A_{m-i}[σᵢ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
These may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
If this process yields at least one full application $r~a₁ \dots aₘ$, we can add a new hypothesis of type $T = B[x₁ ↦ a₁, \dots, xₘ ↦ aₘ]$, meaning we generate the subgoal $Γ, h : T$ (unless $Γ$ already contains a hypothesis of type $T$).

@@

\paragraph{Forward State.}
We associate to each goal in Aesop's search tree a forward state that reflects the goal's context.
Let $\States$ be the set of forward states, $\Hyps$ the set of hypotheses, $\Rules$ the set of rules, $\CMatches$ the set of complete matches and $\powerset(X)$ the powerset of $X$.
The forward state's interface then consists of two functions
\begin{equation*}
  \addHyp : \States → \Hyps → \States × \powerset(\Rules × \CMatches) \qquad \delHyp : \States → \Hyps → \States
\end{equation*}
The $\addHyp$ function adds a hypothesis to the forward state.
Hence, if $s$ is the forward state corresponding to context $Γ$ and $\addHyp(s, h : T) = (s', C)$, then $s'$ is the forward state corresponding to $Γ,\, h : T$.
Similarly, if $s$ is the forward state corresponding to $Γ, h : T$, then $\delHyp(s, h : T)$ is the forward state corresponding to $Γ$.
The second output of $\addHyp$, $C$, is a set of pairs $(r, m)$ where $r$ is a rule and $m$ is a complete match for $r$.
These are the matches that were completed by adding $h : T$.

@@

When a hypothesis is added to a rule state, two things can happen.
First, the hypothesis may allow us to extend partial matches already present in the state.
For instance, consider the rule $r : \All{x}{P~x → Q~x → R~x → S~x}$ and a rule state containing the partial match $[h₁]$, which assigns some hypothesis $h₁ : P~a$ to the first slot of type $P~x$.
Adding a new hypothesis $h₂ : Q~a$ to this rule state means that we should add to the state the extended match $[h₁, h₂]$.

Second, adding a match may trigger the insertion of further matches.
Suppose that before adding $h₁$ and $h₂$, we had already added a hypothesis $h₀ : R~a$ to the rule state.
Once the partial match $[h₁, h₂]$ arrives, we should recognise that it can be extended with $h₀$ to obtain the (complete) match $[h₁, h₂, h₀]$.

\paragraph{Variable Indices.}
Extending the matches above was possible only because all hypotheses are compatible on the shared variable $x$, i.e.\ they all instantiate $x$ with the same term $a$.
In general, to extend a match $m$ with a hypothesis $h : T$ at slot $i$, we need to ensure that the substitution of $m$, $\sub(m)$, is compatible with the substitution of $h$, $\sub_{i}(h)$.

@@

\begin{algorithm}
  \SetKwInOut{Precondition}{Precondition}
  \SetKwInOut{Postcondition}{Postcondition}
  \Precondition{$H$ and $M$ contain the hypotheses and matches corresponding to a context $Γ$}
  \Input{A hypothesis $h : T$ matching slot $i$ of $r$ with substitution $σ$}
  \Output{Complete matches of $r$ in context $Γ,\, h : T$ that contain $h$}
  \Postcondition{$H$ and $M$ contain the hypotheses and matches corresponding to $Γ,\, h : T$}
  \BlankLine
  \Procedure{$\mathrm{addHyp}(h : T)$}{
    \eIf{$i = 1$}{
      $\mathrm{addMatch}([h])$
    }{
      \For{$x$ \KwIn $\vars(i)$}{
        Insert $h$ into the set $H_{x}(i, σ(h))$
      }
      \KwLet $\mathit{ms} ≔ \bigcap_{x \in \vars(i)} M_{x}(i-1, σ(x))$ \\
      $\mathit{completeMs} \gets ∅$ \\
      \For{$m$ \KwIn $\mathit{ms}$}{
        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ++ [h])$
      }
      \Return{$\mathit{completeMs}$}
    }
  }
  \BlankLine % without these, the caption is not offset properly
  \BlankLine
  \caption{Adding a hypothesis to the rule state for a rule $r$}%
  \label{alg:addhyp}
\end{algorithm}

The rule state's $\addHyp$ method (Algorithm~\ref{alg:addhyp}) adds a hypothesis $h : T$ matching slot $i$ to the rule state.
(If a hypothesis matches multiple slots, $\addHyp$ is called multiple times.)
For $i = 1$, we add the singleton match $[h]$ to the rule state.
For $i > 1$, we first insert $h$ into the hypothesis index $H$.
Then we query the match index $M$ for matches that can be extended by $h$ and add the extended matches to the rule state.
While adding matches, we may discover new complete matches, which are collected and returned.

% TODO pre- and postcondition?
\begin{algorithm}
  \Input{A partial match $m$ of $r$ at level $i$ with substitution $σ$}
  \Output{A set of complete matches of $r$}
  \BlankLine
  \Procedure{$\mathrm{addMatch}(h : T)$}{
    \eIf{$i$ is complete}{
      \Return{$\{m\}$}
    }{
      \For{$x$ \KwIn $\vars(i + 1)$}{
        Insert $m$ into the set $M_{x}(i, σ(x))$
      }
      \KwLet $\mathit{hs} ≔ \bigcap_{x \in \vars(i + 1)} H_{x}(i + 1, σ(x))$ \\
      $\mathit{completeMs} \gets ∅$ \\
      \For{$h$ \KwIn $\mathit{hs}$}{
        $\mathit{completeMs} \gets \mathit{completeMs} ∪ \mathrm{addMatch}(m ++ [h])$
      }
      \Return{$\mathit{completeMs}$}
    }
  }
  \BlankLine % without these, the caption is not offset properly
  \BlankLine
  \caption{Adding a match to the rule state for a rule $r$}%
  \label{alg:addmatch}
\end{algorithm}

@@

\paragraph{Variable Clusters.}
We have so far assumed that for each slot $i > 1$, the set of shared variables $\vars(i)$ is nonempty, i.e.\ each slot shares at least one variable with at least one of the previous slots.
This assumption is violated for rules such as $\All{x~y}{P~x → Q~y → R~x~y}$: slot 2 ($Q~y$) shares no variables with slot 1 ($P~x$).
As a result, our above algorithms fail.
Suppose we are adding the hypothesis $h₂ : Q~b$, having already added the hypothesis $h₁ : P~a$ and hence the partial match $[h₁]$.
The set of matches compatible with $h₂$ is supposedly $\bigcap_{x \in \vars(2)} M_{x}(i-1, σ(x)) = ∅$, but the match $[h₁]$ can be extended with $h₂$.
In fact, any match at level 1 can be extended with any hypothesis matching slot 2 since the two slots are independent.

@@

\paragraph{Example 1.}
As a first example, consider a rule state for the transitivity rule $r₁ : \All{x~y~z}{x ≤ y → y ≤ z → x ≤ z}$ to which we add, in order, the hypotheses
$h₁ : 301 ≤ 302$ and $h₂ : 300 ≤ 301$.
The rule has two slots of types $x ≤ y$ and $y ≤ z$ which share the variable $y$.

The hypothesis $h₁ : 301 ≤ 302$ unifies with both slots, so we run Algorithm~\ref{alg:addhyp} twice.
For slot 1, unification yields the substitution $\{x ↦ 301, y ↦ 302\}$.
Since this is the first slot, we start a new partial match $[h₁]$, which Algorithm~\ref{alg:addmatch} inserts into the set $M_{y}(1, 302)$.
This match cannot be extended since the hypothesis index $H$ does not contain a hypothesis for slot 2.

For slot 2, unification yields the substitution $σ ≔ \{y ↦ 301, z ↦ 302\}$, so we insert $h₁$ into $H_{y}(2, 301)$.
No matches at level 1 can be extended by $h₁$ since
\[
  \bigcap_{x \in \vars(2)} M_{x}(1, σ(x)) = M_{y}(1, 301) = \emptyset
\]
At this point, $H$ and $M$ are in the following state.
\begin{center}
\begin{tikzpicture}[outer sep=auto, level distance=12mm, scale=0.50]
  \node[xshift = 0] {$H_{y}$}
    child {node {2}
      child {node {$301$}
        child {node {$\{h₁\}$}}}
    %  child {node {$300$}
    %    child {node {$\{\}$}}}
    };
  \node[xshift=160] {$M_y$}
    child {node {1}
      child {node {$302$}
        child {node {$\{[h₁]\}$}}}
    %  child {node {$301$}
    %    child {node {$\{\}$}}}
    };
    %child {node[xshift=20] {2}
    %  child {node {$301$}
    %    child {node[yshift=-2em] {$\{\}$}}}};
\end{tikzpicture}
\end{center}

Now we add $h₂ : 300 ≤ 301$, which again unifies with both slots.
For slot 1, $\sub₁(h₂)(y) = 301$, so we add the match $[h₂]$ to $M_{y}(1, 301)$.
We then check for hypotheses in slot 2 that also instantiate $y$ with $301$, and indeed we find $M_{y}(2, 301) = \{h₁\}$.
Hence we return the complete match $[h₂, h₁]$, which induces a new hypothesis $h₃ : 300 ≤ 302$ to be added later.

@@

We now consider a rule $r₂ : \All{x~y}{A~x~y → B~x~y → C~x~y}$ with multiple shared variables.
We add hypotheses $h₁ : B~a~a$, $h₂ : A~a~b$ and $h₃ : B~a~b$ to the rule state of $r₂$, starting with $h₁$:

@@

We now add $h_2 : A~a~b$, which initially generates a match $m = [h₂]$ that is added to $M$:

@@

\paragraph{Match Equivalence.}
Naively, two matches $[h₁, \dots, hₙ]$ and $[h₁', \dots, hₙ']$ would be considered equal if they are structurally equal, i.e.\ $h₁ = h₁'$, \dots, $hₙ = hₙ'$.
However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
The complete matches $m₁ ≔ [h₁, h₃]$ and $m₁' ≔ [h₁, h₄]$ are structurally different, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
Generating both $m₁$ and $m₁'$ would therefore be redundant.
Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete $m₂$ and $m₂'$.
Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each variable $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.
To ensure that we respect definitional equality, the instantiations in the substitutions are kept in RPINF.

@@

Given a set of propositions $\mathcal{P} := \{P_{i,j} | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i} | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : P_{i,1}\to \dots P_{i,6} \to Q_i$.
The procedures are then run on a context containing exactly $\mathcal{P}$.
In this case, we sketch the behaviour of the stateless algorithm:
It pick a first hypothesis, $h_{i,j} : P_{i,j}$.
If $j < 6$, the index returns nothing and it selects another hypothesis.
For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
This gives a complexity of about $O(n^2)$.
